---
title: "Herpetofauna and Historic Climate"
author: "Alisa Hernandez"
date: "6/29/2021"
output: ioslides_presentation
---

```{r logistics, include=FALSE}
#########################################################################
#Title: Herpetofauna Community and Climate of Cabrillo National Monument
#Purpose: To understand and visualize the relationship of herpetofauna community and climate factors within Cabrillo National Monument 1995-2020.
#Author: Alisa Hernandez
#Created: 6/29/21
#Edited: 6/29/21
#########################################################################
library(readr)
library(readxl)
library(extrafont)
library(PNWColors)
library(viridis)
library(lubridate)
library(broom)
library(rstatix)
library(vegan)
library(tidyverse)
library(patchwork)
library(mgcv)


##Importing Data sets - pitfall array data and climate data 1995-2020
pitfall <- read_excel("../AH_Herps/CABR_pitfall_data.xlsx")
  
SD_climate<-read_csv("../AH_Herps/SD_Intl_Airport_1995-2021.csv")

####Cleaning data
#Selecting only the columns of interest, fixing column names for matching and simplicity
pitfall_limited<-pitfall%>%
  group_by(SurveyYear, Season, Month, Date1, Type, SPECIES)%>%
  rename(Species=SPECIES, SurveyDate=Date1)%>%
  summarise(Count=n())
  
#Filtering for herpetofauna community
herps<-filter(pitfall_limited, Type %in% c('Lizard', 'LIZARD', 'Snake', 'SNAKE'))

#Converting to usable dates, making precipitation values into numeric form, selecting columns of interest, summarizing hourly data into daily averages
climate<-SD_climate%>%
  mutate(SurveyYear=year(date), Month=month(date), SurveyDate=date(date), hourlyprecipitation=as.numeric(hourlyprecipitation))%>%
  group_by(SurveyDate, SurveyYear, Month)%>%
  summarise(dailydewpt=mean(na.omit(hourlydewpointtemperature)),
            dailydrybulb=mean(na.omit(hourlydrybulbtemperature)),
            dailyrelhum=mean(na.omit(hourlyrelativehumidity)),
            dailyprecip = mean(na.omit(hourlyprecipitation)))
       
     
####Joining Data
herps_climate<-herps%>%
  left_join(climate, by = c("SurveyDate", "SurveyYear", "Month"))


####Wide format transition
##Herpetofauna and climate
herps_climate_wide <- herps_climate %>%
  group_by(SurveyYear, Month, Season, Species) %>%
  select(-c(Type, SurveyDate))%>%
  pivot_wider(names_from = Species, values_from = Count, values_fill = 0)

#Herps wide format
herps_wide <- herps %>%
  select(-c(Type, SurveyDate))%>%
  pivot_wider(names_from = Species, values_from = Count, values_fill = 0)

#Climate longer format for CCA, compiles climate factors into 2 columns.
climate_longer <- climate%>%
  pivot_longer(cols=starts_with("daily"),
               names_to = 'weather_type', values_to = 'weather_val')
#Reordering so that the first column is the numeric climate values.
climate_longer2<-climate_longer[, c(5,4,1,2,3)]

#Climate wide format
climate_wide<-climate_longer%>%
  select(-c(1:3))%>%
  pivot_wider(names_from = weather_type, values_from = weather_val)

#Color palette for figures
pal=pnw_palette("Bay", 4, type = "discrete")

my_theme<- theme_classic()+ 
  theme(plot.title = element_text(family="Calibri", size = 20),
    axis.title = element_text(size=15, family = "Calibri"),
                                  axis.text = element_text(size=12, color="black", family = "Calibri"))
```


## The questions I want to answer
1. How has San Diego climate changed in the last twenty five years?
2. How has the Cabrillo National Monument herpetofauna community changed in the last 25 years?
3. Has the herpetofauna community been changed by climate?
  + Which species have been the most successful?
4. Can we predict which species will be successful under climate change conditions?



## Question 1 - 25 years of San Diego climate

```{r climate, }
dewpt<- ggplot(climate,
       aes(x=SurveyDate,
           y=dailydewpt))+
  geom_point()+
  geom_smooth(method=lm, se=FALSE)+
   theme_classic()+
   scale_fill_manual(values=pal)+
  ggtitle("Daily dewpoint over time")

dewpt_mod <- lm(dailydewpt ~ SurveyDate, data = climate)
summary(dewpt_mod) ##p-value 2.918e-11, residuals 0.004469

precip<- ggplot(climate, aes(x=SurveyYear,y=dailyprecip))+
   geom_col(color="light blue", fill="light blue")+
   theme_classic()+
   scale_fill_manual(values = pnw_palette("Sailboat",1))+
   ggtitle("Daily precipitation over time")

######Included figures
 #Temperature 
temp<- ggplot(climate, aes(x=SurveyDate,y=dailydrybulb))+
   geom_point(color = "orange")+
   my_theme+
   ggtitle("Daily air temperature over time")+
  labs(y="Air Temperature (\u00B0F)", x="Year")

ggsave("Daily air temp.png")
 
drybulb_mod <- lm(dailydrybulb ~ SurveyDate, data = climate)
summary(drybulb_mod) ##p-value <2.2e-16, residuals 0.016
 
 
hum<- ggplot(climate, aes(x=SurveyDate,y=dailyrelhum))+
   geom_point(color = "light blue")+
   my_theme+
   ggtitle("Daily relative humidity over time")+
  labs(y="Percent Humidity (%)", x="Year")

ggsave("Daily humidity.png")
 
 hum_mod<-lm(dailyrelhum ~ SurveyDate, data=climate)
 summary(hum_mod) ## p-value <2.2e-16, adjusted r-squared 0.05044
 

 
 (temp + hum)
 ggsave("Daily temp and hum.png")
```

## Question 2 - 25 years of herpetofauna community

```{r herp community, }
ggplot(herps, aes(x=SurveyYear, y=Count, fill=Species))+
  geom_col()+
  theme_classic()+
  ggtitle("Herpetofauna community over time")+
  ggsave("Herpetofauna community over time.png")

ggplot(herps%>%filter(Type=="Lizard"|Type=="LIZARD"), aes(x=SurveyYear, y=Count, fill=Species))+
  geom_col()+
  theme_classic()+
  ggtitle("Lizard community over time")

ggplot(herps%>%filter(Type=="Snake"|Type=="SNAKE"), aes(x=SurveyYear, y=Count, fill=Species))+
  geom_col()+
  scale_color_viridis()+
  theme_classic()+
  ggtitle("Snake community over time")

##################################
##################################
##################################
ungroup(herps) %>%
filter(Species=='SCOC'|Species=='UTST'|Species=='CNHY'|Species=='MALA') %>%
group_by(SurveyYear, Species) %>%
summarise(Count = sum(Count)) %>%
ungroup() %>%
mutate(Species2 = if_else(Species == 'SCOC', 'W. Fence Lizard',
                 if_else(Species == 'UTST', 'Side Blotch',
                 if_else(Species == 'CNHY', 'O.T. Whiptail',
                         'Rattlesnake')))) %>%
ggplot(aes(x=SurveyYear, y= Count, color = Species2))+
  geom_point()+
  geom_line() + 
  facet_wrap(~Species2)+ 
  my_theme+
  ggtitle("Herpetofauna Focus Species Over Time")+
  labs(y="Species Abundance", x="Year") +
  theme(legend.position = 'none')
  
ggsave("Herpetofauna Focus Species Over Time.png")
```

## Question 3 - Climate and community analysis

```{r pressure}
#Attempted on-parametric multiple regression
#########They aren't working :( why?

herps_climate_NMDS <- metaMDS(herps_climate_wide[,-c(1:3)], k = 2, trymax = 100)
#error I keep getting: Error in if (autotransform && xam > 50) { : missing value where TRUE/FALSE needed


herps_cca = cca(herps_wide[,-c(1:3)], climate_longer2[,-c(3:5)])
#error I keep getting: Error in rowSums(X) : 'x' must be numeric

#Trying a single lm...
lm_totalherp_count.season<-lm(Count~Season, data=herps_climate)

summary(totalherpcount.precip_LM)

ggplot(herps_climate,
       aes(x=dailyprecip,
           y=Count))+
  geom_point()+
  geom_smooth(method="lm")

##Precipitation lm for top 4 species
lm_UTST_precip<-lm(Count~dailyprecip, data=herps_climate%>%filter(Species=="UTST"))
summary(lm_UTST_precip) ### p-value: 0.1129

lm_SCOC_precip<-lm(Count~dailyprecip, data=herps_climate%>%filter(Species=="SCOC"))
summary(lm_SCOC_precip) ### p-value: 0.3271

lm_CNHY_precip<-lm(Count~dailyprecip, data=herps_climate%>%filter(Species=="CNHY"))
summary(lm_CNHY_precip) ### p-value: 0.4281

lm_MALA_precip<-lm(Count~dailyprecip, data=herps_climate%>%filter(Species=="MALA"))
summary(lm_MALA_precip) ### p-value: 0.5636

##Dew point lm for top 4 species
lm_UTST_dewpt<-lm(Count~dailydewpt, data=herps_climate%>%filter(Species=="UTST"))
summary(lm_UTST_dewpt) ### p-value: 0.0005111
ggplot(herps_climate%>%filter(Species=="UTST"), aes(x=dailydewpt, y=Count))+
  geom_point()+
  geom_smooth(method="lm")

lm_SCOC_dewpt<-lm(Count~dailydewpt, data=herps_climate%>%filter(Species=="SCOC"))
summary(lm_SCOC_dewpt) ### p-value: 3.128e-08
ggplot(herps_climate%>%filter(Species=="SCOC"), aes(x=dailydewpt, y=count))+
  geom_point()+
  geom_smooth(method="lm")

lm_CNHY_dewpt<-lm(Count~dailydewpt, data=herps_climate%>%filter(Species=="CNHY"))
summary(lm_CNHY_dewpt) ### p-value: 0.8032

lm_MALA_dewpt<-lm(Count~dailydewpt, data=herps_climate%>%filter(Species=="MALA"))
summary(lm_MALA_dewpt) ### p-value: 0.1914

##Dry bulb lm for top 4 species
lm_UTST_drybulb<-lm(Count~dailydrybulb, data=herps_climate%>%filter(Species=="UTST"))
summary(lm_UTST_drybulb) ### p-value: 6.049e-11
ggplot(herps_climate%>%filter(Species=="UTST"), aes(x=dailydrybulb, y=Count))+
  geom_point()+
  geom_smooth(method="lm")

lm_SCOC_drybulb<-lm(Count~dailydrybulb, data=herps_climate%>%filter(Species=="SCOC"))
summary(lm_SCOC_drybulb) ### p-value: 0.1435

lm_CNHY_drybulb<-lm(Count~dailydrybulb, data=herps_climate%>%filter(Species=="CNHY"))
summary(lm_CNHY_drybulb) ### p-value: 0.5703

lm_MALA_drybulb<-lm(Count~dailydrybulb, data=herps_climate%>%filter(Species=="MALA"))
summary(lm_MALA_drybulb) ### p-value: 0.3878

##Relative humidity lm for top 4 species
lm_UTST_relhum<-lm(Count~dailyrelhum, data=herps_climate%>%filter(Species=="UTST"))
summary(lm_UTST_relhum) ### p-value: 0.2788

lm_SCOC_relhum<-lm(Count~dailyrelhum, data=herps_climate%>%filter(Species=="SCOC"))
summary(lm_SCOC_relhum) ### p-value: 3.765e-11
ggplot(herps_climate%>%filter(Species=="SCOC"), aes(x=dailyrelhum, y=Count))+
  geom_point()+
  geom_smooth(method="lm")

lm_CNHY_relhum<-lm(Count~dailyrelhum, data=herps_climate%>%filter(Species=="CNHY"))
summary(lm_CNHY_relhum) ### p-value: 0.2415

lm_MALA_relhum<-lm(Count~dailyrelhum, data=herps_climate%>%filter(Species=="MALA"))
summary(lm_MALA_relhum) ### p-value: 0.24


#Calculating richness, now what do I do with it? Attach it to the other data frame and plot it over time?

herps_rich <- tibble('richness' =  specnumber(herps_wide[,-c(1:5)]))

herps_rich <- cbind(herps_rich, herps_wide)

```

``` {R, }
# step 1 - get data
UTST <- herps_climate %>%
  filter(Species == 'UTST') %>%
  # summarize to get abundance across factors (year, temp variance, hum) 
  # hypotheses - lizards related to temp variance, year & humidity
  group_by(SurveyYear) %>%
  summarise(Count = sum(Count),
            Temp_var = var(dailydrybulb),
            Hum = mean(dailyrelhum))

# step 2 - run model
UTST_mod <- glm(Count ~ SurveyYear + Temp_var + Hum, data = UTST)
summary(UTST_mod)

plot(UTST_mod)

# step 3 - plot results
ggplot(data = UTST, 
       mapping = aes(x = Hum, y = Count)) + 
  geom_point() + 
  geom_smooth(method = 'lm')


##Alisa running glm
CNHY <- herps_climate%>%
  filter(Species=='CNHY')%>%
  group_by(SurveyYear) %>%
  summarise(Count = sum(Count),
            Temp_var = var(dailydrybulb),
            Hum = mean(dailyrelhum))

CNHY_mod <- glm(Count ~ SurveyYear + Temp_var + Hum, data = CNHY)
summary(CNHY_mod) ##Pull R2 value from lm, 

ggplot(data = CNHY, 
       mapping = aes(x = Hum, y = Count)) + 
  geom_point() + 
  geom_smooth(method = 'lm')

SCOC <- herps_climate%>%
  filter(Species=='SCOC')%>%
  group_by(SurveyYear) %>%
  summarise(Count = sum(Count),
            Temp_var = var(dailydrybulb),
            Hum = mean(dailyrelhum))

SCOC_mod <- glm(Count ~ SurveyYear + Temp_var + Hum, data = SCOC)
summary(SCOC_mod) ##None significant?

ggplot(data = SCOC, 
       mapping = aes(x = Hum, y = Count)) + 
  geom_point() + 
  geom_smooth(method = 'lm')

MALA <- herps_climate%>%
  filter(Species=='MALA')%>%
  group_by(SurveyYear) %>%
  summarise(Count = sum(Count),
            Temp_var = var(dailydrybulb),
            Hum = mean(dailyrelhum))

MALA2 <- herps_climate%>%
  filter(Species=='MALA')%>%
  group_by(SurveyYear) %>%
  summarise(Count = sum(Count),
            Temp_mean = mean(dailydrybulb),
            Hum = mean(dailyrelhum))

MALA_mod <- glm(Count ~ SurveyYear + Temp_mean + Hum, data = MALA2)
summary(MALA_mod) ##None significant?

ggplot(data = MALA, 
       mapping = aes(x = Hum, y = Count)) + 
  geom_point() + 
  geom_smooth(method = 'lm')

####################################################################
#####################################################################
##################################################################
herps_wide2 <- herps %>%
  group_by(SurveyYear, Species) %>%
  summarise(Count = sum(Count)) %>%
  pivot_wider(names_from = Species, values_from = Count, values_fill = 0)


herps_rich <- tibble('richness' =  specnumber(herps_wide2[,-c(1)]))

herps_rich <- cbind(herps_rich, herps_wide2) %>%
  select(richness, SurveyYear)

herps_rich <- left_join(herps_rich, select(SCOC, -c(Count)), by = 'SurveyYear')


rich_mod<- glm(richness ~ SurveyYear + Temp_var + Hum, data =herps_rich)

summary(rich_mod)
```